Joint estimation of ancestral states and phylogenetic model parameters is inconsistent

### 2017-11-28 joint_inf.tex
rough cut abstract and consistent wording

\begin{abstract}
TODO
\end{abstract}

@@

Classical maximum-likelihood (ML) estimation in phylogenetics operates by integrating out possible ancestral states at the internal nodes of the tree.
Recently, \cite{Sagulenko2017-jo} have suggested using an approximation to ML inference in which the likelihood is maximized jointly across model parameters and ancestral sequences.
This is attractive from a computational perspective: such joint ML inference can proceed according to an iterative procedure in which ML ancestral sequences are first inferred and then model parameters are optimized conditional on the ancestral sequences.
This latter conditional optimization is simpler and more computationally efficient than marginalizing over ancestral sequences and then performing optimization.

Existing consistency proofs for maximum-likelihood phylogenetics \cite{RoyChoudhury2015-ta} apply only to estimating model parameters when the ancestral sequences have been integrated out of the likelihood.
These proofs do not readily extend to include estimating ancestral states.
Moreover, examples of inconsistency arising from including a large number of additional parameters \cite{Neyman1948-tt} indicate that joint inference of trees and ancestral states may not enjoy good statistical properties.
Although \cite{Sagulenko2017-jo} explicitly warn that their approximation is for the case of branch lengths $\ll 1$ and limit themselves to optimizing branch lengths using this method, their work motivates understanding when exactly this approximation breaks down, and what happens when optimizing branch lengths and tree topology using this approximation.

>>>>

\begin{abstract}
Maximum-likelihood phylogenetics is an approach to estimate the phylogenetic parameters of topology and branch length given observed character data.
Since a phylogeny has unknown intermediate states, computing a maximum likelihood solution requires dealing with these latent variables.
Classically, the method of marginal likelihood is employed: the unknown intermediate states typically come from a small set, and so averaging over their finite space allows for consistent estimation of the parameters of interest.
Recently, a computationally-efficient approach has been proposed to maximize over these unknown states as opposed to averaging over them.
While this method of maximum likelihood has the potential to obtain estimates more efficiently, its properties are not well-understood.
We show that, like other methods of profile likelihood where the number of unknowns grows with observations, this method of jointly estimating phylogenetic parameters along with ancestral states is not consistent in general.
There are a number of cases when generating under a four-taxon tree results in a joint estimation procedure converging on the wrong topology.
If the topology is known a priori, there are similar cases where the branch lengths are consistently overestimated.
While these results do not show joint inference is always inconsistent, they indicate further study into where care can be taken when using this approach.
\end{abstract}

@@

Classical maximum likelihood (ML) estimation in phylogenetics operates by integrating out possible ancestral states at the internal nodes of the tree.
Recently, \cite{Sagulenko2017-jo} have suggested using an approximation to ML inference in which the likelihood is maximized jointly across model parameters and ancestral sequences.
This is attractive from a computational perspective: such joint ML inference can proceed according to an iterative procedure in which ML ancestral sequences are first inferred and then model parameters are optimized conditional on the ancestral sequences.
This latter conditional optimization is simpler and more computationally efficient than marginalizing over ancestral sequences and then performing optimization.

Existing consistency proofs for maximum likelihood phylogenetics \cite{RoyChoudhury2015-ta} apply only to estimating model parameters when the ancestral sequences have been integrated out of the likelihood.
These proofs do not readily extend to include estimating ancestral states.
Moreover, examples of inconsistency arising from including a large number of additional parameters \cite{Neyman1948-tt} indicate that joint inference of trees and ancestral states may not enjoy good statistical properties.
Although \cite{Sagulenko2017-jo} explicitly warn that their approximation is for the case of branch lengths $\ll 1$ and limit themselves to optimizing branch lengths using this method, their work motivates understanding when exactly this approximation breaks down, and what happens when optimizing branch lengths and tree topology using this approximation.
